\sys: Property-Guided Program Regeneration\\{\vspace{1em}\large Technical University of Denmark, MSc Thesis {\\\vspace{-1em}} Advisors: Christian Gram Kahlauge and Nikos Vasilakis

### 2025-07-24 docs/paper/draft.tex
Explain query DSL in more detail
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

\usepackage{setspace}
\usepackage{enumerate}
\usepackage{algorithm2e}
\usepackage{algpseudocode}
\usepackage{graphics}
\usepackage{xparse} 
\usepackage{xspace}
\usepackage{multirow}
\usepackage{csvsimple}
\usepackage{balance}
\usepackage[newfloat]{minted}
\usepackage{xcolor}
\usepackage{nicefrac}
\usepackage{siunitx}
\usepackage{array,framed}
\usepackage{booktabs}
\usepackage{color}
\usepackage{soul}
\usepackage{float}
\usepackage{epsfig}
\usepackage{wrapfig}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{adjustbox}
\usepackage{csquotes}
\usepackage{cleveref}
\usepackage{dirtytalk}
\usepackage{textgreek}
\usepackage{oplotsymbl}
\usepackage{listings}
\usepackage{flushend}
\usepackage{float}

@@

\section{Property Language and Extraction}
\label{sec:dsl}

@@

\begin{listing}[t]
  \begin{minted}[frame=lines, fontsize=\small]{prolog}
% --- Base program facts ---
language(p, c).

% --- Plugin: len ---
can(len(p)). 
% Plugin: signature ---
can(signature(p)).
% ---  Plugin: tr ---
can(tr(p, haskell)).
  do(language(p, haskell)) :- do(tr(p, haskell)).
% ---- Axioms ----
:- X, not can(X). % All transformations must be possible
:- language(p, L1), language(p, L2)),
    L1 != L2. % A program can have one language

% --- User query ---
#min len(p). % User query
\end{minted}
  \caption{\textbf{Example \sys logic program.}
  This is a lightly simplified version of the logic program that produces the plan for the idiomatization task in \cref{sec:example}. 
  The program contains facts coming from (1) the input program (written in C),
  (2) three plugins (\ttt{len}, \ttt{signature}, and \ttt{tr}),
  (3) axioms from the logic engine (\eg a program can have one language, an analysis will be applied only if possible, all goals must be satisfied), and
  (4) the user query (the user wants to minimize the program's length).
  }
  \label{lst:logic-example}
\end{listing}

@@

Additionally, the system takes advantage of ASP's least stable model semantics.
Essentially, the planner tries to give the synthesizer the least amount of information
possible, as long as all regeneration goals are fulfilled.

@@

\heading{Compilation to ASP}
Given a source program written in the \sys query language, the compiler performs a syntactic
de-sugaring into ASP rules~(\cref{fig:asp-compilation}).
These transformations yield a groundable ASP program that can be evaluated by the backend ASP solver.

\begin{figure}[t]
\centering
\small
\begin{tabular}{l}
\textbf{toasp} (q : Query) : ASPCode {$\leftarrow$} \textbf{match} q \textbf{with} \\[0.5ex]
\quad \textbar~$X$: ungroundable($X$) {$\rightarrow$} \ttt{do(}$X$\ttt{)} \\
\quad \textbar~\#\ttt{min}~$X$ {$\rightarrow$} \ttt{do\_min(}$X$\ttt{)} \\
\quad \textbar~\#\ttt{max}~$X$ {$\rightarrow$} \ttt{do\_max(}$X$\ttt{)} \\
\textbf{end}
\end{tabular}
\caption{DSL-to-ASP compilation rules. Ungroundable atoms are wrapped in \ttt{do} to denote candidate actions. Optimization directives are compiled to \ttt{do\_min} and \ttt{do\_max}.}
\label{fig:asp-compilation}
\end{figure}

% TODO: Show graph showing how properties go from the abstract logic domain to the real domain through a translation barrier
\heading{Logic and concrete space}
\sys operates at the boundary between two representational domains: a symbolic
\emph{logic space}, where transformations over program properties are expressed, and
a \emph{concrete space}, where actual programs, inputs, outputs, and traces
reside.

>>>>

\usepackage{setspace}
\usepackage{enumerate}
\usepackage{algorithm2e}
\usepackage{algpseudocode}
\usepackage{graphics}
\usepackage{xparse} 
\usepackage{xspace}
\usepackage{multirow}
\usepackage{csvsimple}
\usepackage{balance}
\usepackage[newfloat]{minted}
\usepackage{xcolor}
\usepackage{nicefrac}
\usepackage{siunitx}
\usepackage{array,framed}
\usepackage{booktabs}
\usepackage{color}
\usepackage{soul}
\usepackage{float}
\usepackage{epsfig}
\usepackage{wrapfig}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{adjustbox}
\usepackage{csquotes}
\usepackage{cleveref}
\usepackage{dirtytalk}
\usepackage{textgreek}
\usepackage{oplotsymbl}
\usepackage{listings}
\usepackage{flushend}
\usepackage{float}
\usepackage{algorithm2e}

@@

\section{Query Language}
\label{sec:dsl}

@@

\begin{listing}[t]
  \begin{minted}[frame=lines, fontsize=\small]{prolog}
% Original query (q.pl)
graph(p_, Ih, Oh) :- graph(p, Ijs, Ojs),
    tr(Ijs, Ih),
    tr(Ojs, Oh).
language(p_, haskell).

% Compiled program
% Base program facts
program(p).
program(p_).
language(p, javascript).
% -----------
% Plugin-provided facts
can(graph(P)) :- language(P, javascript).
can(tr).
% ------------
% Query predicates
can(graph(P, graph_p_Ijs, graph_p_Ojs)) :-
    do(graph(P)),
    program(P).
% ...compilation of tr predicates...
can(language(P, haskell)) :-
    program(P).
% ------------
% Axioms
:- do(language(P, L1)), language(P, _). 
% ----------
% Regeneration plan choice rule
0 { do(X) } 1 :- can(X).
% User query goals
:- not do(graph(p_,tr_graph_p_Ijs_Ih,tr_graph_p_Ojs_Oh)).
:- not do(language(p_, haskell)).
#show do/1.
\end{minted}
  \caption{\textbf{Compiled \sys logic program.}
  The logic program that produces the plan for the idiomatization task in \cref{sec:example}. 
  The program contains facts coming from 
  (1) the input program (written in C),
  (2) three plugins (\ttt{len}, \ttt{fsignt}, and \ttt{tr}),
  (3) axioms from the logic engine (\eg a program can have one language, an analysis will be applied only if possible, all goals must be satisfied), and
  (4) the user query (the user wants to minimize the program's length).
  }
  \label{lst:logic-example}
\end{listing}

@@

Additionally, the system takes advantage of ASP's least stable model semantics.
Intuitively, the planner tries to give the synthesizer the least amount of information
possible, as long as all regeneration goals are fulfilled.

@@

\heading{Compilation to ASP}
Given a source program written in the \sys query language, the compiler
performs a syntactic de-sugaring into ASP rules~(\cref{fig:asp-compilation}).
For each rule $H \leftarrow B_1,\dots,B_n$, the compiler first performs a
pseudo-grounding step that replaces every capitalized variable $V_i$ with a
fresh, unique symbol $\mathit{sym}(r,P,V_i)$, generated from the rule
name $r$, target program $P$, and original variable name $V_i$, proceeding
left-to-right to preserve data dependencies.
Each one of these concrete values will correspond to extracted program properties.
After renaming, the head and body
are emitted as an ASP capability rule of the form $\mathit{can}(H') \leftarrow
\mathit{do}(B_1'),\dots,\mathit{do}(B_n')$. 
All implicit query
goals (rule heads $H$) are compiled into hard constraints $\neg \mathit{do}(H) \rightarrow
\bot$, meaning that the ASP program will be unsatisfiable unless all query goals are met.
Optimization directives $\#\min X$ and $\#\max X$ are mapped to
$\mathit{do_min}(X)$ and $\mathit{do_max}(X)$, respectively.
Finally, \sys
appends program-agnostic axioms and adds a choice rules that allows the solver to select which properties to extract from the program in order to satisfy the query.
This consists as the core of the \sys logic program, and the reason the rule to \ttt{can}/\ttt{do} predicates is required.
The result is the compilation is a valid ASP program which is then evaluated by an ASP solver.
The next section describes how the set of produced predicates is used to construct the regeneration plan.

\begin{figure}[t]
\centering
\small
\begin{tabular}{l}
\textbf{toasp} (q : Query) : ASPCode {$\leftarrow$} \textbf{match} q \textbf{with} \\[0.5ex]
  \quad \textbar~$p$: \textbf{program name} {$\rightarrow$} \ttt{program}$(p)$. \\[0.5ex]
  \quad \textbar~$X$: predicate {$\rightarrow$} \ttt{can}$(X)$ \\[0.5ex]
\quad \textbar~$H \leftarrow B_1,\dots,B_n$
    {$\rightarrow$} 
    $V_i$ 
    $\mapsto$ 
    $\mathit{sym}(r,P,V_i) = r\ttt{\_}P\ttt{\_}V_i$ \\
  \quad \textbar~$H \leftarrow B_1,\dots,B_n$ {$\rightarrow$} \ttt{can}$(H)$\ttt{ :- do}$(B_1)$, \dots, \ttt{do}$(B_n)$. \\[0.5ex]
  \quad \textbar~$H \leftarrow B$  {$\rightarrow$} \ttt{:- not do}$(H)$. \\[0.5ex]
  \quad \textbar~\#\ttt{min}~$X$ {$\rightarrow$} \ttt{do\_min}$(X)$. \\[0.5ex]
  \quad \textbar~\#\ttt{max}~$X$ {$\rightarrow$} \ttt{do\_max}$(X)$. \\[0.5ex]
\textbf{end} \\[1ex]
\multicolumn{1}{p{0.9\linewidth}}{\textit{Axioms and choice rules:}} \\[0.5ex]
\quad \ttt{0 \{ do(X) \} 1 :- can(X).} \\[0.5ex]
\quad \ttt{:- do(language(P,L1)), language(P,\_).} \\
\multicolumn{1}{p{0.9\linewidth}}{\textit{Plugin-supplied rules and facts.}} \\[0.5ex]
\end{tabular}
\caption{\sys query compilation into an ASP program.}
\label{fig:asp-compilation}
\end{figure}

\heading{Logic program resolution}
Once the input query is compiled into an ASP program, \sys invokes an
off-the-shelf ASP solver to compute a stable model that satisfies the query.
The model consists of a set of selected $\mathit{do}$ atoms representing the regeneration plan.
\sys then executes this plan~\cref{alg:resolution}.
First, \sys discards all plugin-activation atoms from the plan, 
as they are only relevant to the logic engine.
Then \sys iterates through each
$\mathit{do}(f(a_1,\dots,a_n))$ atom.
If an argument value is missing, the plugin responsible for producing
it (found by parsing the variable's name) is executed and produced program properties are stored in a value map.
Once all actions have been executed in dependency order, the extracted properties are passed to the regeneration phase.

\begin{algorithm}[t]
\caption{Resolution of a regeneration plan from an ASP model into concrete program-extracted values.}
\label{alg:resolution}
\DontPrintSemicolon
\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\Input{ASP model (set of \(\mathit{do}\) atoms)}
\Output{Map $V \mapsto V_p$ of concrete values for each property \(V\) in the model}
% initialize
\ttt{discard} all plugin‐activation atoms from \(\mathit{plan}\)\;
\(\mathit{values}\;\gets\;\emptyset\)\;
% main loop
\ForEach{\(d \in \mathit{plan}\)}{
  \If{\(d = \mathit{do}(f(a_1,\dots,a_n))\)}{
    \For{each argument \(a_i\) of \(f\)}{
      \If{\(a_i \notin \mathit{values}\)}{
        invoke plugin responsible for \(a_i\)\;
        \(\mathit{values}[a_i] \gets\) result of plugin\;
      }
    }
    invoke plugin \(f\) with \((a_1,\dots,a_n)\)\;
  }
}
\end{algorithm}

\heading{Logic and concrete space}
\sys operates at the boundary between two representational domains: a symbolic
\emph{logic space}, where transformations over program properties are expressed, and
a \emph{concrete space}, where actual programs, inputs, outputs, and traces
reside.
